---
title: "Duke Blue Devils — Team Season Grade Report"
author: "Pat Rankin"
format:
  html:
    toc: true
    toc-title: "On This Page"
    echo: false
editor: visual
execute:
  message: false
  warning: false
---

```{r}
#| label: packages
library(hoopR)
library(dplyr)
library(readr)
library(readxl)
library(ggplot2)
library(tidyr)
library(plotly)
```

```{r}
#| label: settings
my_team      <- "Duke"
season_year  <- 2026
season_label <- "2025-26"
```

```{r}
#| label: ingest
#| include: false

team_box <- hoopR::load_mbb_team_box(seasons = season_year)

team_data <- team_box %>%
  filter(team_location == my_team |
           team_name == my_team |
           team_short_display_name == my_team)

if (nrow(team_data) == 0) {
  stop("No games found for '", my_team, "'.")
}

team_name_full <- team_data$team_display_name[1]
games_played   <- nrow(team_data)
wins           <- sum(team_data$team_winner == TRUE,  na.rm = TRUE)
losses         <- sum(team_data$team_winner == FALSE, na.rm = TRUE)
win_pct        <- wins / games_played * 100
```

```{r}
#| label: averages
#| include: false

avg <- team_data %>%
  mutate(
    team_score    = as.numeric(team_score),
    opp_score     = as.numeric(opponent_team_score),
    fg_pct        = as.numeric(field_goal_pct),
    three_pct     = as.numeric(three_point_field_goal_pct),
    ft_pct        = as.numeric(free_throw_pct),
    total_reb     = as.numeric(total_rebounds),
    ast           = as.numeric(assists),
    stl           = as.numeric(steals),
    blk           = as.numeric(blocks),
    tov           = as.numeric(turnovers),
    pt_diff       = team_score - opp_score
  ) %>%
  summarise(
    pts       = mean(team_score, na.rm = TRUE),
    opp_pts   = mean(opp_score,  na.rm = TRUE),
    fg_pct    = mean(fg_pct,     na.rm = TRUE),
    three_pct = mean(three_pct,  na.rm = TRUE),
    ft_pct    = mean(ft_pct,     na.rm = TRUE),
    reb       = mean(total_reb,  na.rm = TRUE),
    ast       = mean(ast,        na.rm = TRUE),
    stl       = mean(stl,        na.rm = TRUE),
    blk       = mean(blk,        na.rm = TRUE),
    tov       = mean(tov,        na.rm = TRUE),
    pt_diff   = mean(pt_diff,    na.rm = TRUE)
  )
```

```{r}
#| label: grading
#| include: false

score_to_letter <- function(score) {
  if      (score >= 4.0) "A+"
  else if (score >= 3.7) "A"
  else if (score >= 3.3) "A-"
  else if (score >= 3.0) "B+"
  else if (score >= 2.7) "B"
  else if (score >= 2.3) "B-"
  else if (score >= 2.0) "C+"
  else if (score >= 1.7) "C"
  else if (score >= 1.3) "C-"
  else if (score >= 1.0) "D+"
  else if (score >= 0.7) "D"
  else if (score >= 0.3) "D-"
  else                   "F"
}

grade_stat <- function(value, thresholds, lower_is_better = FALSE) {
  if (lower_is_better) {
    if      (value <= thresholds["A"]) gpa <- 4.0
    else if (value <= thresholds["B"]) gpa <- 3.0
    else if (value <= thresholds["C"]) gpa <- 2.0
    else if (value <= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  } else {
    if      (value >= thresholds["A"]) gpa <- 4.0
    else if (value >= thresholds["B"]) gpa <- 3.0
    else if (value >= thresholds["C"]) gpa <- 2.0
    else if (value >= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  }
  score_to_letter(gpa)
}

letter_to_gpa <- function(l) {
  switch(l,
    "A+"=4.3,"A"=4.0,"A-"=3.7,
    "B+"=3.3,"B"=3.0,"B-"=2.7,
    "C+"=2.3,"C"=2.0,"C-"=1.7,
    "D+"=1.3,"D"=1.0,"D-"=0.7,"F"=0.0)
}

# Benchmarks: NCAA Division I per-game averages (2024-25)
grade_scoring    <- grade_stat(avg$pts,       c(A=80, B=74, C=68, D=62))
grade_defense    <- grade_stat(avg$opp_pts,   c(A=62, B=67, C=72, D=77), lower_is_better=TRUE)
grade_fg         <- grade_stat(avg$fg_pct,    c(A=48, B=45, C=42, D=39))
grade_three      <- grade_stat(avg$three_pct, c(A=37, B=34, C=31, D=28))
grade_ft         <- grade_stat(avg$ft_pct,    c(A=75, B=70, C=65, D=60))
grade_rebounding <- grade_stat(avg$reb,       c(A=40, B=37, C=34, D=31))
grade_assists    <- grade_stat(avg$ast,       c(A=16, B=14, C=12, D=10))
grade_turnovers  <- grade_stat(avg$tov,       c(A=11, B=13, C=15, D=17), lower_is_better=TRUE)
grade_win_pct    <- grade_stat(win_pct,       c(A=70, B=60, C=50, D=40))

all_grades  <- c(grade_scoring, grade_defense, grade_fg, grade_three,
                 grade_ft, grade_rebounding, grade_assists,
                 grade_turnovers, grade_win_pct)
weights <- c(0.15, 0.15, 0.12, 0.10, 0.08, 0.10, 0.10, 0.10, 0.10)

overall_gpa   <- sum(sapply(all_grades, letter_to_gpa) * weights)
overall_grade <- score_to_letter(overall_gpa)

grade_df <- data.frame(
  Category = c("Scoring", "Defense", "FG%", "3-Point%",
               "Free Throw%", "Rebounding", "Assists",
               "Ball Security", "Win %"),
  Value    = c(avg$pts, avg$opp_pts, avg$fg_pct, avg$three_pct,
               avg$ft_pct, avg$reb, avg$ast, avg$tov, win_pct),
  Grade    = all_grades,
  GPA      = sapply(all_grades, letter_to_gpa)
)

grade_df$Grade_Group <- case_when(
  grade_df$GPA >= 3.7 ~ "A",
  grade_df$GPA >= 2.7 ~ "B",
  grade_df$GPA >= 1.7 ~ "C",
  grade_df$GPA >= 0.7 ~ "D",
  TRUE                ~ "F"
)
```

## `r team_name_full` — `r season_label` Season Report

**Record:** `r wins`-`r losses` (`r sprintf("%.1f", win_pct)`% win rate) \| **Games Graded:** `r games_played`

**Overall Season Grade: `r overall_grade`**

------------------------------------------------------------------------

## Category Grades

```{r}
#| label: grade-chart

grade_colors <- c("A"="#2ecc71","B"="#3498db","C"="#f39c12","D"="#e67e22","F"="#e74c3c")

ggplot(grade_df, aes(x = reorder(Category, GPA), y = GPA, fill = Grade_Group)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Grade), hjust = -0.2, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = grade_colors, name = "Grade") +
  scale_y_continuous(limits = c(0, 5.2), breaks = c(0,1,2,3,4)) +
  coord_flip() +
  labs(
    title    = paste(team_name_full, "— Category Grades"),
    subtitle = paste(season_label, "NCAA Season"),
    x        = NULL,
    y        = "Grade Points (4.0 scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```

## Scoring Trend

```{r}
#| label: scoring-trend

trend_data <- team_data %>%
  mutate(
    game_date  = as.Date(game_date),
    team_score = as.numeric(team_score),
    opp_score  = as.numeric(opponent_team_score),
    result     = ifelse(team_winner == TRUE, "Win", "Loss")
  ) %>%
  arrange(game_date) %>%
  mutate(game_num = row_number())

ggplot(trend_data, aes(x = game_num)) +
  geom_line(aes(y = team_score), color = "#2ecc71", linewidth = 1) +
  geom_line(aes(y = opp_score),  color = "#e74c3c", linewidth = 1, linetype = "dashed") +
  geom_point(aes(y = team_score, color = result), size = 2.5) +
  scale_color_manual(values = c("Win"="#2ecc71","Loss"="#e74c3c"), name = "Result") +
  geom_hline(yintercept = avg$pts, linetype = "dotted", color = "#2ecc71", linewidth = 0.7) +
  labs(
    title    = paste(team_name_full, "— Scoring by Game"),
    subtitle = "Green = team score  |  Red dashed = opponent score  |  Dotted = season average",
    x        = "Game Number",
    y        = "Points"
  ) +
  theme_minimal(base_size = 13)
```

## Key Stats Summary

```{r}
#| label: stats-table

summary_tbl <- data.frame(
  Statistic        = c("Points Per Game", "Opp. Points Per Game", "Point Differential",
                       "Field Goal %", "Three-Point %", "Free Throw %",
                       "Rebounds Per Game", "Assists Per Game",
                       "Steals Per Game", "Blocks Per Game", "Turnovers Per Game"),
  Value            = c(
    sprintf("%.1f", avg$pts),
    sprintf("%.1f", avg$opp_pts),
    sprintf("%+.1f", avg$pt_diff),
    sprintf("%.1f%%", avg$fg_pct),
    sprintf("%.1f%%", avg$three_pct),
    sprintf("%.1f%%", avg$ft_pct),
    sprintf("%.1f", avg$reb),
    sprintf("%.1f", avg$ast),
    sprintf("%.1f", avg$stl),
    sprintf("%.1f", avg$blk),
    sprintf("%.1f", avg$tov)
  )
)

knitr::kable(summary_tbl, align = c("l","r"), col.names = c("Statistic","Season Avg"))
```

## How the Grade Is Calculated

Each category is graded against NCAA Division I per-game averages and converted to a 4.0 scale. The overall grade is a weighted average of all categories, with scoring and defense carrying the most weight (15% each). Ball security and win percentage each account for 10%.

The grade scale is: **A+** (elite) · **A / A-** (excellent) · **B** range (above average) · **C** range (average) · **D** range (below average) · **F** (well below average).

------------------------------------------------------------------------

## Logistic Regression: Can Field Goal % Predict a Win?

```{r}
#| label: read-duke-xlsx
team_Duke_2026 <- read_excel("../../../pipeline/ncaa/basketball/data/team_stats/team_Duke_2026.xlsx")
```

### Does field goal percentage predict whether Duke wins or loses?

```{r}
#| label: fg-pct-jitter
team_Duke_2026 |>
  mutate(result = if_else(team_winner, "win", "loss")) |>
  ggplot(aes(x = field_goal_pct, y = if_else(team_winner, 0, 0.5), color = result)) +
  geom_jitter(height = 0.2, width = 0.2, alpha = 0.3, size = 4) +
  scale_color_manual(values = c("win" = "#00539b", "loss" = "#c0392b")) +
  labs(
    title = "Can Field Goal Percentage Predict Winning?",
    x = "Field Goal %"
  ) +
  scale_y_continuous(breaks = c(0, 0.5), labels = c("Win", "Lose")) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.line.x = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5),
    axis.title.y = element_blank()
  )
```

> **Interpretation:** Based on the actual numbers from Duke's data (35 wins, 3 losses) — average FG% is 49.2% in wins vs. 49.7% in losses, essentially identical. Wins span 38–62% FG while losses span 45–52% FG, meaning the losses sit comfortably inside the range of shooting performances Duke also *won* with. With only 3 losses in 38 games, there isn't enough variation in outcomes for field goal percentage to emerge as a meaningful predictor — Duke wins so consistently that shooting efficiency alone doesn't distinguish their few losses from their many wins.

```{r}
#| label: model-fg-pct
model <- glm(team_winner ~ field_goal_pct,
             data = team_Duke_2026,
             family = binomial)

summary(model)
```

The coefficient on field goal percentage is negative and the p-value (0.890) is high, confirming FG% does not play a statistically significant role in whether Duke wins.

### Adding a Second Predictor: Turnovers

```{r}
#| label: model-fg-tov
model <- glm(team_winner ~ field_goal_pct + turnovers,
             family = binomial, data = team_Duke_2026)

summary(model)
```

Field goal percentage is still not a relevant variable in the model. Turnovers has a positive coefficient, suggesting it may influence winning, but its p-value (0.996) is far too high to be statistically significant, and the AIC (26.971) increased versus the single-predictor model.

### Turnovers, Three-Point %, and Rebounds

```{r}
#| label: model-tov-3pt-reb
model <- glm(team_winner ~ turnovers + three_point_field_goal_pct +
               total_rebounds, data = team_Duke_2026, family = binomial)

summary(model)
```

Turnovers and total rebounds are the more meaningful variables in this model. Total rebounds has a low p-value, indicating statistical significance. The AIC is lower than the prior models — a better trade-off between fit and complexity — and the residual deviance is also lower, indicating a better fit to the data.

### Probability of Winning by Turnover Rate

```{r}
#| label: turnover-probability-curve
model <- glm(team_winner ~ turnovers, data = team_Duke_2026, family = binomial)

turnover_range <- data.frame(turnovers = seq(min(team_Duke_2026$turnovers, na.rm = TRUE),
                                              max(team_Duke_2026$turnovers, na.rm = TRUE),
                                              length.out = 500))

turnover_range$prob <- predict(model, newdata = turnover_range, type = "response")

plot_ly(data = turnover_range,
        x = ~turnovers,
        y = ~prob,
        type = "scatter",
        mode = "lines",
        name = "Probability Curve") |>
  layout(
    title = "Probability of Team Winning by Turnover Rate",
    xaxis = list(title = "Turnover Rate"),
    yaxis = list(title = "Probability of Winning")
  )
```

**Bottom line:** for this Duke team, field goal percentage doesn't meaningfully predict whether they win or lose — their few losses look like they came from something other than poor shooting (e.g. a tougher opponent, an off night elsewhere). Turnovers and rebounding carry more predictive signal, though with only 3 losses on the season, every model here is working with very little variation to explain.

------------------------------------------------------------------------

## Interactive Stat Distribution Explorer

Use the dropdown below to see how any of Duke's key stats were distributed across the season — every game becomes one entry in the histogram.

```{r}
#| label: histogram-explorer-setup

hist_vars <- c(
  "Points Scored"    = "team_score",
  "Field Goal %"     = "field_goal_pct",
  "Three-Point %"    = "three_point_field_goal_pct",
  "Free Throw %"     = "free_throw_pct",
  "Total Rebounds"   = "total_rebounds",
  "Assists"          = "assists",
  "Turnovers"        = "turnovers",
  "Steals"           = "steals",
  "Blocks"           = "blocks"
)
```

```{r}
#| label: histogram-explorer

p <- plot_ly()

for (i in seq_along(hist_vars)) {
  var   <- hist_vars[[i]]
  label <- names(hist_vars)[i]
  p <- p |> add_histogram(
    x       = team_Duke_2026[[var]],
    name    = label,
    marker  = list(color = "#00539b"),
    visible = (i == 1)
  )
}

buttons <- lapply(seq_along(hist_vars), function(i) {
  visible <- rep(FALSE, length(hist_vars))
  visible[i] <- TRUE
  list(
    method = "update",
    args   = list(
      list(visible = visible),
      list(title = paste("Distribution of", names(hist_vars)[i]),
           xaxis  = list(title = names(hist_vars)[i]))
    ),
    label = names(hist_vars)[i]
  )
})

p |> layout(
  title  = paste("Distribution of", names(hist_vars)[1]),
  xaxis  = list(title = names(hist_vars)[1]),
  yaxis  = list(title = "Number of Games"),
  updatemenus = list(
    list(
      type    = "dropdown",
      active  = 0,
      x = 1, xanchor = "right", y = 1.15, yanchor = "top",
      buttons = buttons
    )
  )
)
```

> Prefer to explore this in R directly? The same variable-selector-and-histogram idea is available as a standalone Shiny app: [`app.R`](../../../pipeline/ncaa/basketball/data/team_stats/app.R). Open it in RStudio and click **Run App** for a version with an adjustable bin-count slider.
