---
title: "Shai Gilgeous-Alexander: Logistic Regression Example"
subtitle: "Does his individual box score predict whether the Thunder win?"
author: "Pat Rankin"
editor: visual
execute:
  message: false
  warning: false
---

### Setup

```{r}
#| label: packages

library(tidyverse)
library(broom)
library(plotly)
```

```{r}
#| label: ingest

sga <- read_csv(
  "../../pipeline/nba/data/player_stats/nba_player_Shai_Gilgeous-Alexander_2025.csv",
  show_col_types = FALSE
) |>
  filter(team_name == "Thunder") |>  # drop the All-Star Weekend "Team Chuck" rows
  mutate(
    field_goal_pct = field_goals_made / field_goals_attempted * 100,
    team_winner    = as.logical(team_winner)
  )

nrow(sga)
```

Filtering to `team_name == "Thunder"` matters here — the raw file also contains two 2025 All-Star Game rows (`team_name == "Team Chuck"`), which aren't real Thunder games and would contaminate `team_winner`.

### Single-predictor logistic regression

**Question: does the number of points Shai scores in a game predict whether the Thunder win that game?**

```{r}
#| label: single-predictor

model_pts <- glm(team_winner ~ points, data = sga, family = binomial)
summary(model_pts)

pts_range <- data.frame(
  points = seq(min(sga$points, na.rm = TRUE), max(sga$points, na.rm = TRUE), length.out = 500)
)
pts_range$prob <- predict(model_pts, newdata = pts_range, type = "response")

plot_ly(data = pts_range, x = ~points, y = ~prob, type = "scatter", mode = "lines") |>
  layout(
    title = "Probability of a Thunder Win by Shai's Points Scored",
    xaxis = list(title = "Points"),
    yaxis = list(title = "Probability of Winning")
  )
```

#### Interpretation

`points` is a significant predictor (p = 0.032), with a positive coefficient (0.081, odds ratio ≈ 1.084) — each additional point he scores multiplies the Thunder's odds of winning that game by about 1.08, roughly an 8% increase in odds per point. That's a real but modest relationship: scoring is one ingredient in a team win, not the whole recipe, which is exactly what you'd expect from a single player's point total predicting a team outcome.

### Adding more predictors

**Question: once we account for shooting efficiency, assists, and turnovers together, which of Shai's box score stats actually move the needle on winning?**

```{r}
#| label: multi-predictor

model1 <- glm(team_winner ~ points + field_goal_pct + assists + turnovers,
              data = sga, family = binomial)
summary(model1)
```

#### Interpretation

Once `field_goal_pct`, `assists`, and `turnovers` are in the model together, `points` (p = 0.21) and `field_goal_pct` (p = 0.54) both drop out of significance — their earlier apparent effect was largely riding on top of other stats. The standout is `turnovers` (p = 0.030, odds ratio ≈ 0.71): each additional turnover Shai commits multiplies the Thunder's odds of winning by about 0.71, roughly a 29% drop in odds per turnover. `assists` (p = 0.56) isn't significant.

*Ball security* is the stat most reliably tied to a Thunder win in Shai's box scores — not volume scoring or shooting percentage.

### Visualizing the multi-predictor model

```{r}
#| label: odds-ratio-plot

coef_df <- tidy(model1, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)")

ggplot(coef_df, aes(x = estimate, y = reorder(term, estimate))) +
  geom_point(size = 3, color = "#007AC1") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2, color = "#007AC1") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  scale_x_log10() +
  labs(x = "Odds ratio (log scale)", y = NULL,
       title = "Effect of each Shai Gilgeous-Alexander box score stat on odds of a Thunder win") +
  theme_minimal()
```

### Model discrimination: AUC

```{r}
#| label: auc

complete_idx <- as.numeric(rownames(model.frame(model1)))
preds  <- fitted(model1)
actual <- sga$team_winner[complete_idx]
wins   <- preds[actual == TRUE]
losses <- preds[actual == FALSE]
auc <- mean(outer(wins, losses, ">")) + 0.5 * mean(outer(wins, losses, "=="))
cat("AUC:", auc, "\n")
```

#### Interpretation

AUC ≈ 0.68. That means if you picked one Thunder win and one Thunder loss at random from Shai's games and used only his personal box score (points, FG%, assists, turnovers) to guess which was the win, the model would guess right about 68% of the time — clearly better than a coin flip, but far from perfect. That makes sense: one player's stat line is informative but incomplete — it can't fully explain a team outcome that also depends on four teammates and the opponent.

### Takeaways from this example

-   **Turnovers, not scoring, is the most statistically reliable driver of a Thunder win in Shai's box score** — the opposite of what you might guess from his gaudy scoring averages.
-   **Points alone looks predictive in isolation** (p = 0.032) but loses significance once turnovers and assists are controlled for — a caution against reading too much into a single-predictor model.
-   **A single player's box score only partially predicts a team win** (AUC ≈ 0.68) — a sanity check that matches basketball intuition: no one player's line fully determines a team result.
