---
title: "NCAA Team Season Grade Report"
author: "Pat Rankin"
format: html
editor: visual
execute:
  message: false
  warning: false
code-fold: true
---

```{r}
#| label: packages
library(hoopR)
library(dplyr)
library(readr)
library(tidyverse)
library(tidyr)
```

## ⚙️ Your Settings — Change These Before Rendering

```{r}
#| label: settings

# ============================================================
#  FILL IN YOUR INFO BELOW — this is the only section
#  you need to change each time you run this report.
# ============================================================

# Which team do you want to grade?
# Use the school's common name exactly as ESPN lists it.
# Examples: "Duke", "Kansas", "Houston", "Gonzaga", "Alabama State"
my_team <- "Duke"

# Which season? Use the year the season ENDS.
# Example: the 2025-26 season is written as 2026.
season_year <- 2026

# Season label shown in the report (for display only)
season_label <- "2025-26"

# ============================================================
#  THAT'S IT — scroll down to render and read your report.
# ============================================================
```

```{r}
#| label: ingest
#| include: false

# Pull the full season from hoopR and filter to your team
team_box <- hoopR::load_mbb_team_box(seasons = season_year)

team_data <- team_box %>%
  filter(team_location == my_team |
           team_name == my_team |
           team_short_display_name == my_team)

if (nrow(team_data) == 0) {
  stop("No games found for '", my_team, "'. Check spelling or run: ",
       "team_box %>% distinct(team_location) %>% arrange(team_location) %>% print(n=400)")
}

team_name_full <- team_data$team_display_name[1]
games_played   <- nrow(team_data)
wins           <- sum(team_data$team_winner == TRUE,  na.rm = TRUE)
losses         <- sum(team_data$team_winner == FALSE, na.rm = TRUE)
win_pct        <- wins / games_played * 100
```

```{r}
#| label: averages
#| include: false

avg <- team_data %>%
  mutate(
    team_score    = as.numeric(team_score),
    opp_score     = as.numeric(opponent_team_score),
    fg_pct        = as.numeric(field_goal_pct)             * 100,
    three_pct     = as.numeric(three_point_field_goal_pct) * 100,
    ft_pct        = as.numeric(free_throw_pct)             * 100,
    total_reb     = as.numeric(total_rebounds),
    ast           = as.numeric(assists),
    stl           = as.numeric(steals),
    blk           = as.numeric(blocks),
    tov           = as.numeric(turnovers),
    pt_diff       = team_score - opp_score
  ) %>%
  summarise(
    pts       = mean(team_score, na.rm = TRUE),
    opp_pts   = mean(opp_score,  na.rm = TRUE),
    fg_pct    = mean(fg_pct,     na.rm = TRUE),
    three_pct = mean(three_pct,  na.rm = TRUE),
    ft_pct    = mean(ft_pct,     na.rm = TRUE),
    reb       = mean(total_reb,  na.rm = TRUE),
    ast       = mean(ast,        na.rm = TRUE),
    stl       = mean(stl,        na.rm = TRUE),
    blk       = mean(blk,        na.rm = TRUE),
    tov       = mean(tov,        na.rm = TRUE),
    pt_diff   = mean(pt_diff,    na.rm = TRUE)
  )
```

```{r}
#| label: grading
#| include: false

# --- helpers ---
score_to_letter <- function(score) {
  if      (score >= 4.0) "A+"
  else if (score >= 3.7) "A"
  else if (score >= 3.3) "A-"
  else if (score >= 3.0) "B+"
  else if (score >= 2.7) "B"
  else if (score >= 2.3) "B-"
  else if (score >= 2.0) "C+"
  else if (score >= 1.7) "C"
  else if (score >= 1.3) "C-"
  else if (score >= 1.0) "D+"
  else if (score >= 0.7) "D"
  else if (score >= 0.3) "D-"
  else                   "F"
}

grade_stat <- function(value, thresholds, lower_is_better = FALSE) {
  if (lower_is_better) {
    if      (value <= thresholds["A"]) gpa <- 4.0
    else if (value <= thresholds["B"]) gpa <- 3.0
    else if (value <= thresholds["C"]) gpa <- 2.0
    else if (value <= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  } else {
    if      (value >= thresholds["A"]) gpa <- 4.0
    else if (value >= thresholds["B"]) gpa <- 3.0
    else if (value >= thresholds["C"]) gpa <- 2.0
    else if (value >= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  }
  score_to_letter(gpa)
}

letter_to_gpa <- function(l) {
  switch(l,
    "A+"=4.3,"A"=4.0,"A-"=3.7,
    "B+"=3.3,"B"=3.0,"B-"=2.7,
    "C+"=2.3,"C"=2.0,"C-"=1.7,
    "D+"=1.3,"D"=1.0,"D-"=0.7,"F"=0.0)
}

# --- grade each category ---
# Benchmarks: NCAA Division I per-game averages (2024-25)
grade_scoring    <- grade_stat(avg$pts,       c(A=80, B=74, C=68, D=62))
grade_defense    <- grade_stat(avg$opp_pts,   c(A=62, B=67, C=72, D=77), lower_is_better=TRUE)
grade_fg         <- grade_stat(avg$fg_pct,    c(A=48, B=45, C=42, D=39))
grade_three      <- grade_stat(avg$three_pct, c(A=37, B=34, C=31, D=28))
grade_ft         <- grade_stat(avg$ft_pct,    c(A=75, B=70, C=65, D=60))
grade_rebounding <- grade_stat(avg$reb,       c(A=40, B=37, C=34, D=31))
grade_assists    <- grade_stat(avg$ast,       c(A=16, B=14, C=12, D=10))
grade_turnovers  <- grade_stat(avg$tov,       c(A=11, B=13, C=15, D=17), lower_is_better=TRUE)
grade_win_pct    <- grade_stat(win_pct,       c(A=70, B=60, C=50, D=40))

all_grades  <- c(grade_scoring, grade_defense, grade_fg, grade_three,
                 grade_ft, grade_rebounding, grade_assists,
                 grade_turnovers, grade_win_pct)
weights <- c(0.15, 0.15, 0.12, 0.10, 0.08, 0.10, 0.10, 0.10, 0.10)

overall_gpa   <- sum(sapply(all_grades, letter_to_gpa) * weights)
overall_grade <- score_to_letter(overall_gpa)

# --- grade table for charts ---
grade_df <- data.frame(
  Category = c("Scoring", "Defense", "FG%", "3-Point%",
               "Free Throw%", "Rebounding", "Assists",
               "Ball Security", "Win %"),
  Value    = c(avg$pts, avg$opp_pts, avg$fg_pct, avg$three_pct,
               avg$ft_pct, avg$reb, avg$ast, avg$tov, win_pct),
  Grade    = all_grades,
  GPA      = sapply(all_grades, letter_to_gpa)
)

grade_df$Grade_Group <- case_when(
  grade_df$GPA >= 3.7 ~ "A",
  grade_df$GPA >= 2.7 ~ "B",
  grade_df$GPA >= 1.7 ~ "C",
  grade_df$GPA >= 0.7 ~ "D",
  TRUE                ~ "F"
)
```

## `r team_name_full` — `r season_label` Season Report

**Record:** `r wins`-`r losses` (`r sprintf("%.1f", win_pct)`% win rate) \| **Games Graded:** `r games_played`

**Overall Season Grade: `r overall_grade`**

------------------------------------------------------------------------

## Category Grades

```{r}
#| label: grade-chart

grade_colors <- c("A"="#2ecc71","B"="#3498db","C"="#f39c12","D"="#e67e22","F"="#e74c3c")

ggplot(grade_df, aes(x = reorder(Category, GPA), y = GPA, fill = Grade_Group)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Grade), hjust = -0.2, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = grade_colors, name = "Grade") +
  scale_y_continuous(limits = c(0, 5.2), breaks = c(0,1,2,3,4)) +
  coord_flip() +
  labs(
    title    = paste(team_name_full, "— Category Grades"),
    subtitle = paste(season_label, "NCAA Season"),
    x        = NULL,
    y        = "Grade Points (4.0 scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```

## Scoring Trend

```{r}
#| label: scoring-trend

trend_data <- team_data %>%
  mutate(
    game_date  = as.Date(game_date),
    team_score = as.numeric(team_score),
    opp_score  = as.numeric(opponent_team_score),
    result     = ifelse(team_winner == TRUE, "Win", "Loss")
  ) %>%
  arrange(game_date) %>%
  mutate(game_num = row_number())

ggplot(trend_data, aes(x = game_num)) +
  geom_line(aes(y = team_score), color = "#2ecc71", linewidth = 1) +
  geom_line(aes(y = opp_score),  color = "#e74c3c", linewidth = 1, linetype = "dashed") +
  geom_point(aes(y = team_score, color = result), size = 2.5) +
  scale_color_manual(values = c("Win"="#2ecc71","Loss"="#e74c3c"), name = "Result") +
  geom_hline(yintercept = avg$pts, linetype = "dotted", color = "#2ecc71", linewidth = 0.7) +
  labs(
    title    = paste(team_name_full, "— Scoring by Game"),
    subtitle = "Green = team score  |  Red dashed = opponent score  |  Dotted = season average",
    x        = "Game Number",
    y        = "Points"
  ) +
  theme_minimal(base_size = 13)
```

## Key Stats Summary

```{r}
#| label: stats-table

summary_tbl <- data.frame(
  Statistic        = c("Points Per Game", "Opp. Points Per Game", "Point Differential",
                       "Field Goal %", "Three-Point %", "Free Throw %",
                       "Rebounds Per Game", "Assists Per Game",
                       "Steals Per Game", "Blocks Per Game", "Turnovers Per Game"),
  Value            = c(
    sprintf("%.1f", avg$pts),
    sprintf("%.1f", avg$opp_pts),
    sprintf("%+.1f", avg$pt_diff),
    sprintf("%.1f%%", avg$fg_pct),
    sprintf("%.1f%%", avg$three_pct),
    sprintf("%.1f%%", avg$ft_pct),
    sprintf("%.1f", avg$reb),
    sprintf("%.1f", avg$ast),
    sprintf("%.1f", avg$stl),
    sprintf("%.1f", avg$blk),
    sprintf("%.1f", avg$tov)
  )
)

knitr::kable(summary_tbl, align = c("l","r"), col.names = c("Statistic","Season Avg"))
```

## How the Grade Is Calculated

Each category is graded against NCAA Division I per-game averages and converted to a 4.0 scale. The overall grade is a weighted average of all categories, with scoring and defense carrying the most weight (15% each). Ball security and win percentage each account for 10%.

The grade scale is: **A+** (elite) · **A / A-** (excellent) · **B** range (above average) · **C** range (average) · **D** range (below average) · **F** (well below average).

### Logistic Regression model

```{r}
# Read in data
library(readxl)
team_Duke_2026 <- read_excel("C:/Users/Pat/Desktop/website2/pipeline/ncaa/basketball/data/team_stats/team_Duke_2026.xlsx")
```

### Use single numerical categorical(predictor) to see if field goal percentage can predict winning games? Can field goal percentage predict whether a team wins or not?

```{r}
team_Duke_2026 |>
  mutate(result = if_else(team_winner, "win", "loss")) |>
  ggplot(aes(x = field_goal_pct, y = if_else(team_winner, 0, 0.5), color = result)) +
  geom_jitter(height = 0.2, width = 0.2, alpha = 0.3, size = 4) +
  scale_color_manual(values = c("win" = "blue", "loss" = "red")) +
  labs(
    title = "Can field goal percentage predict winning",
    x = "Field Goal %"
  ) +
  scale_y_continuous(breaks = c(0, 0.5), labels = c("Win", "Lose")) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.line.x = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5),
    axis.title.y = element_blank()
  )
```

### interpretation

Based on the actual numbers from Duke's data (35 wins, 3 losses):

-   **Average FG%:** 49.2% in wins vs. 49.7% in losses — essentially identical, and if anything slightly *higher* in the losses.

-   **Range:** wins span 38–62% FG, losses span 45–52% FG — the losses sit comfortably inside the range of shooting performances Duke also *won* with.

**What the plot shows:** the blue (win) and red (loss) dots are scattered across the same stretch of the x-axis with no clean separation. There's no point on the FG% scale where you can draw a line and say "left of this, they lose; right of this, they win." The red dots aren't clustered at the low end — they're mixed in among the blue ones.

**Why:** with only 3 losses in 38 games, there just isn't enough variation in outcomes for field goal percentage to emerge as a predictor — Duke wins so consistently that shooting efficiency alone doesn't distinguish their few losses from their many wins. That matches the logistic regression result you should see if you run `summary(model)` on `glm(team_winner ~ field_goal_pct, ...)`: the coefficient is small and not statistically significant (p ≈ 0.89).

**Bottom line:** for this Duke team, field goal percentage doesn't meaningfully predict whether they win or lose — their losses look like they came from something other than poor shooting (e.g. a tougher opponent, an off night elsewhere). in the summary the coefficient is negative which tells us that field goal percege does not play a significant role in terms of Duke winning or not. Also the p value(0.890) is high.

```{r}
model <- glm(team_winner ~ field_goal_pct,
             data = team_Duke_2026,
             family = binomial)

summary(model)

```

### Adding a numerical variable to the model

```{r}
model <- glm(team_winner ~ field_goal_pct + turnovers, 
             family = binomial, data = team_Duke_2026)

summary(model)
```

### Interpretation of model

Field goal percentage is still not a relevant variable in the model, however turnovers has a positive coefficient which indicates that it does influence winning games, however the p value associated with turnovers is high(0.996) which tells us it is not statistically significant. The AIC score of 26.971 has also increased from the earlier model.

```{r}
model <- glm(team_winner ~ turnovers + three_point_field_goal_pct +
               total_rebounds, data = team_Duke_2026, family = binomial)
               
summary(model)
```

### Interpretation 

Turnovers and total rebounds are meaningful variables to have in the model. Total rebounds have a low p value which indicates statistical signifcance. The AIC score is lower which suggests a better trade-off between fit and complexity. The residual deviance also was lower which indicates how well the model fits the data

### Logistic Model Regression

```{r}
library(plotly)

model <- glm(team_winner ~ turnovers, data = team_Duke_2026, family = binomial)

turnover_range <- data.frame(turnovers = seq(min(team_Duke_2026$turnovers,
                                              na.rm = TRUE),
                                          max(team_Duke_2026$turnovers, 
                                              na.rm = TRUE),
                                          length.out = 500))

turnover_range$prob <- predict(model, newdata = turnover_range, type = "response")

plot_ly(data = turnover_range,
        x = ~turnovers,
        y = ~prob,
        type ="scatter",
        mode = "lines",
        name = "Probability Curve") |> 
  
layout(
  title = "Probability of Team Winning by Turnover Rate",
  xaxis = list(title = "Turnover Rate"),
  yaxis = list(title = "Probability of Winning"))
  

                                          
                                          

                                          




```
