---
title: "Chiefs 2024: A Season Built on Close Wins"
subtitle: "How Kansas City turned one-score games into a 15-2 record"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show Code"
execute:
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)

kc <- read_csv("../../pipeline/nfl/data/nfl_team_KC_2024.csv", show_col_types = FALSE) |>
  mutate(
    margin     = team_score - opp_score,
    result     = if_else(won, "Win", "Loss"),
    close_game = abs(margin) <= 3
  ) |>
  arrange(week)

games       <- nrow(kc)
wins        <- sum(kc$won)
losses      <- games - wins
win_pct     <- wins / games * 100
close_wins  <- sum(kc$close_game & kc$won)
close_games <- sum(kc$close_game)

best_win  <- kc |> filter(won)  |> slice_max(margin, n = 1)
worst_loss <- kc |> filter(!won) |> slice_min(margin, n = 1)

home_record <- kc |> filter(home_away == "Home") |> summarise(w = sum(won), n = n())
away_record <- kc |> filter(home_away == "Away") |> summarise(w = sum(won), n = n())
```

The 2024 Kansas City Chiefs didn't blow teams off the field — they beat them late, over and over. Across `r games` games, Kansas City went **`r wins`-`r losses`** (`r sprintf("%.1f", win_pct)`% win rate), and the story of that record is written in the margins: `r close_wins` of their `r wins` wins came by three points or fewer.

## Week-by-Week Scoring

```{r}
#| label: scoring-trend

ggplot(kc, aes(x = week)) +
  geom_line(aes(y = team_score), color = "#E31837", linewidth = 1) +
  geom_line(aes(y = opp_score), color = "#666666", linewidth = 1, linetype = "dashed") +
  geom_point(aes(y = team_score, color = result), size = 3) +
  scale_color_manual(values = c("Win" = "#E31837", "Loss" = "#1a1a1a"), name = "Result") +
  labs(
    title    = "Kansas City Chiefs — Points Scored by Week (2024)",
    subtitle = "Red = points scored | Gray dashed = points allowed",
    x = "Week", y = "Points"
  ) +
  theme_minimal(base_size = 13)
```

Nothing about that chart looks dominant. The Chiefs rarely pulled away — their average scoring margin for the season was just `r sprintf("%+.1f", mean(kc$margin))` points per game, thin for a team that finished 15-2.

## Living on the Margin

```{r}
#| label: margin-chart

ggplot(kc, aes(x = factor(week), y = margin, fill = close_game)) +
  geom_col() +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
  scale_fill_manual(values = c("TRUE" = "#FFB81C", "FALSE" = "#E31837"),
                     labels = c("TRUE" = "Decided by ≤ 3", "FALSE" = "Other"),
                     name = NULL) +
  labs(
    title = "Margin of Victory / Defeat by Week",
    x = "Week", y = "Point Margin (Team − Opponent)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

Five of Kansas City's wins were decided by three points or fewer — a Week 2 escape against Cincinnati (26-25), a two-point win over Denver in Week 10, a three-point win over Carolina, and two-point wins over Las Vegas and the Chargers in consecutive weeks (13 and 14). That's a full quarter of the season's wins coming down to the final possession.

## Home vs. Away

```{r}
#| label: home-away-table

home_away_tbl <- tibble(
  Split   = c("Home", "Away"),
  Record  = c(
    paste0(home_record$w, "-", home_record$n - home_record$w),
    paste0(away_record$w, "-", away_record$n - away_record$w)
  ),
  Games   = c(home_record$n, away_record$n)
)

knitr::kable(home_away_tbl, align = c("l", "c", "c"))
```

Kansas City played it close no matter where the game was — the road record held up nearly as well as the record at home, a sign of a team that didn't lean on a friendly building to win football games.

## The Best and Worst of It

The best result of the year came against **`r best_win$opponent`** in Week `r best_win$week`, a `r best_win$team_score`-`r best_win$opp_score` win (+`r best_win$margin`). The worst came in **Week `r worst_loss$week`** against **`r worst_loss$opponent`**, a `r worst_loss$team_score`-`r worst_loss$opp_score` blowout loss — a Week 18 finale that looks far more like a team resting starters with the top seed already locked up than a real measuring stick.

## Takeaway

Strip out that meaningless Week 18 finale and Kansas City finished the meaningful part of its schedule 15-1, and did it the hard way — five wins by a field goal or less, a road record nearly as good as its home record, and no game all season that looked like a laugher in their favor. Teams that win close games all year either have something real (a quarterback who doesn't blink, a defense that gets one stop when it has to) or they're due for regression. The 2024 Chiefs bet on the former, and for one more year, it held up.
