---
title: "Boston celtics linear regression"
editor: visual
warning = FALSE
messages = FALSE
---

```{r}
install.packages("corrgram")
library(corrgram)
library(corrplot)
library(tidyverse)
library(readxl)
library(gtExtras)

nba_team_Celtics_2026 <- read_excel("team_stats/nba_team_Celtics_2026.xlsx")
View(nba_team_Celtics_2026)
```

### Numeric correlation for linear regression

```{r}
# returns only numeric columns
num.cols <- sapply(nba_team_Celtics_2026,is.numeric)

# filter numerical columns for correlation
cor.data <- cor(nba_team_Celtics_2026[,num.cols])
print(cor.data)
```

```{r}
library(corrplot)

numeric_cols <- c("team_score", "assists", "blocks", "defensive_rebounds",
  "fast_break_points", "field_goal_pct", "field_goals_made",
  "field_goals_attempted", "free_throw_pct", "free_throws_made",
  "free_throws_attempted", "largest_lead", "offensive_rebounds",
  "points_in_paint", "steals", "team_turnovers", "three_point_field_goal_pct",
  "three_point_field_goals_made", "three_point_field_goals_attempted",
  "total_rebounds", "turnovers", "opponent_team_score")

cor.data <- cor(nba_team_Celtics_2026[, numeric_cols], use = "complete.obs")

# clear any stale/ghosted plot first
while (dev.cur() > 1) dev.off()

corrplot(cor.data, method = "color",
         type = "upper",       # drop the redundant mirrored half
         tl.col = "black",
         tl.cex = 0.7,         # shrink label text so long names fit
         tl.srt = 45,          # angle labels instead of vertical stacking
         mar = c(0, 0, 1, 0))
```

### Interpretation

How would I interpret this plot

Reading a `corrplot(method="color")` heatmap:

**The basics**

-   Each cell = correlation between the row variable and column variable, ranging -1 to +1.

-   **Blue = positive** (as one goes up, the other tends to go up). **Red/brown = negative** (as one goes up, the other tends to go down).

-   **Darker/more saturated = stronger relationship.** Pale/near-white Ōēł no linear relationship.

-   The diagonal is always dark blue (=1) ŌĆö a variable perfectly correlates with itself. Ignore it.

-   With `type="upper"` you only see half the matrix (it's symmetric, so the lower half is redundant).

**Applying it to your basketball data** (using the correlations from your earlier `cor()` output):

-   `team_score` vs `field_goals_made` = **0.88** ŌåÆ strong dark blue. Unsurprising: making more shots wins you more points.

-   `team_score` vs `three_point_field_goal_pct` = **0.73**, vs `assists` = **0.70** ŌåÆ also strong blue. Efficient/team-oriented offense tracks with scoring.

-   `defensive_rebounds` vs `total_rebounds` = **0.76** ŌåÆ strong blue, but this is partly definitional (total rebounds includes defensive rebounds).

-   `free_throws_made` vs `free_throws_attempted` = **0.94** ŌåÆ near-perfect blue. This pair (and similar "made" vs "attempted" pairs) is *tautological*, not insightful ŌĆö ignore these, they're just measuring the same underlying event twice.

-   `defensive_rebounds` vs `opponent_team_score` = **-0.46** ŌåÆ moderate red/negative. Makes basketball sense: out-rebounding opponents on defense limits their second-chance points.

-   `steals` vs `defensive_rebounds` = **-0.38** ŌåÆ red. Worth investigating but could just reflect play style differences between teams.

**What to actually look for**

1.  Scan for the darkest cells (blue or red) *excluding* the diagonal and the obvious made/attempted pairs ŌĆö those are your real signal.

2.  Look at the row/column for `team_score` (or whatever your target variable is) specifically ŌĆö that tells you what predicts winning/scoring.

3.  Treat anything pale as noise, not "no effect" ŌĆö correlation only captures linear relationships.

4.  Remember correlation ŌēĀ causation ŌĆö e.g. high assists correlating with score doesn't mean *forcing* more assists causes more points; both could stem from good shot-making.

If you dropped the ID columns (`game_id`, `team_id`, etc.) like I suggested, you shouldn't see any blank/white NA cells anymore ŌĆö if you still do, those are variables with zero variance or all-NA in your filtered data.

```{r}
# corrgram

corrgram(nba_team_Celtics_2026)
```

```{r}
num.cols <- c("team_score", "assists", "blocks", "defensive_rebounds",
  "fast_break_points", "field_goal_pct", "field_goals_made",
  "field_goals_attempted", "flagrant_fouls", "fouls", "free_throw_pct",
  "free_throws_made", "free_throws_attempted", "largest_lead",
  "offensive_rebounds", "points_in_paint", "steals", "team_turnovers",
  "technical_fouls", "three_point_field_goal_pct",
  "three_point_field_goals_made", "three_point_field_goals_attempted",
  "total_rebounds", "total_technical_fouls", "total_turnovers",
  "turnover_points", "turnovers", "opponent_team_score")

cor.data <- cor(nba_team_Celtics_2026[, num.cols], use = "complete.obs")

library(corrplot)
while (dev.cur() > 1) dev.off()
corrplot(cor.data, method = "color", type = "upper",
         tl.col = "black", tl.cex = 0.7, tl.srt = 45, mar = c(0,0,1,0))
```

### Interpretation

Here's how to read that matrix, using the Celtics data you pasted as the working example.

**Structure of the table**

-   Every row and column is a variable; each cell is the **Pearson correlation** between that row/column pair, from **-1 to +1**.

-   It's symmetric ŌĆö the cell at `(team_score, assists)` equals the cell at `(assists, team_score)` ŌĆö so you only need to scan half of it.

-   The diagonal is always `1.0` (a variable perfectly correlates with itself) ŌĆö skip it.

-   `0` = no linear relationship. `+1`/`-1` = perfect positive/negative relationship. Anything below \~0.3 in magnitude is usually too weak to act on.

**What predicts `team_score` (row 6) ŌĆö the strongest signals:**

| **Variable** | **r** | **Read** |
|:---|:---|:---|
| `field_goals_made` | 0.879 | Strongest driver ŌĆö obviously, makes = points |
| `field_goal_pct` | 0.864 | Shooting efficiency tracks scoring closely |
| `three_point_field_goal_pct` | 0.731 | 3pt shooting efficiency matters a lot |
| `assists` | 0.701 | Ball movement ŌåÆ scoring |
| `three_point_field_goals_made` | 0.696 |  |
| `largest_lead` | 0.680 | Games they score more in, they tend to blow out (or vice versa) |
| `fast_break_points` | 0.444 | Moderate ŌĆö transition offense helps but isn't dominant |
| `points_in_paint` | 0.362 | Moderate |

**Near-zero ŌĆö don't bother interpreting these:**\
`free_throw_pct` (0.046), `offensive_rebounds` (-0.075), `total_rebounds` (-0.072) ŌĆö essentially no linear relationship with scoring in this data.

**Negative and worth a second look:**

-   `turnovers` vs `team_score` = **-0.268** ŌĆö sensible, fewer giveaways, more points.

-   `defensive_rebounds` vs `opponent_team_score` = **-0.456**, `total_rebounds` vs `opponent_team_score` = **-0.520** ŌĆö boxing out limits opponent scoring, as expected.

-   `steals` vs `defensive_rebounds` = **-0.381** ŌĆö teams/games with more steals tend to have fewer defensive boards, possibly a style-of-play tradeoff (gambling for steals vs. staying home to box out).

**Ignore these ŌĆö they're tautological, not insight:**

-   `free_throws_made` vs `free_throws_attempted` = 0.942

-   `three_point_field_goal_pct` vs `three_point_field_goals_made` = 0.798

-   `defensive_rebounds` vs `total_rebounds` = 0.763

-   Any "made" vs "attempted" or "X_pct" vs "X_made" pair ŌĆö these are mathematically related by definition, so a high correlation tells you nothing new.

**Caveats to keep in mind:**

1.  This is one team's games in one season ŌĆö sample size is likely \~2 dozen to \~82 rows, so treat weaker correlations (anything under \~0.3ŌĆō0.4) as noise, not real effects.

2.  Correlation ŌēĀ causation ŌĆö e.g., `assists` correlating with `team_score` doesn't prove passing *causes* scoring; both likely stem from playing well overall.

3.  Watch for the "made/attempted/pct" family of variables ŌĆö they're not independent and can crowd out more meaningful patterns visually if left in the matrix.

If your goal is to find what actually *drives* wins/scoring, focus on the non-tautological variables with \|r\| \> 0.4 relative to `team_score` and `opponent_team_score` ŌĆö that's `field_goals_made`, `field_goal_pct`, `three_point_field_goal_pct`, `assists`, `largest_lead`, `defensive_rebounds`/`total_rebounds` (via opponent scoring), and `turnovers`.

```{r}
celtics_num <- nba_team_Celtics_2026[, has_variance]
celtics_num <- celtics_num[, !names(celtics_num) %in% c("game_id", "opponent_team_id")]
```

```{r}
nba_team_Celtics_2026 <- read_excel("C:/Users/Pat/Desktop/website2/pipeline/nba/data/team_stats/nba_team_Celtics_2026.xlsx")
```

```{r}
library(corrgram)

# keep only numeric columns with actual variance
has_variance <- sapply(nba_team_Celtics_2026, function(col) {
  is.numeric(col) && sd(col, na.rm = TRUE) > 0
})

celtics_num <- nba_team_Celtics_2026[, has_variance]

# optional: also drop ID columns that technically vary but aren't real stats
celtics_num <- celtics_num[, !names(celtics_num) %in% c("game_id", "opponent_team_id")]

# confirm no NAs remain
sum(is.na(celtics_num))   # should print 0

corrgram(celtics_num, order = TRUE,
         lower.panel = panel.shade,
         upper.panel = panel.pie,
         text.panel = panel.txt)
```

```{r}
png("celtics_corrgram.png", width = 1800, height = 1800, res = 180)
corrgram(celtics_num, order = TRUE,
         lower.panel = panel.shade,
         upper.panel = panel.pie,
         text.panel = panel.txt)
dev.off()
```

### Logistic Regression Model

Does field goal percentage(predictor variable) have any influece on whether the team wins or not

```{r}

library(plotly)

nba_team_Celtics_2026 <- read_excel("nba_team_Celtics_2026.xlsx")
View(nba_team_Celtics_2026)
names(nba_team_Celtics_2026)
```

```{r}
model <- glm()
```
